We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-07 paper/paper.tex
new refs

\begin{abstract}
We use numerical simulations to build an explicit probability
distribution for the asphericity in the satellite distribution around
galaxies similar to the Local Group (LG) in the Lambda Cold Dark
Matter (LCDM) paradigm. 
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of resolution, simulated physics and criteria to select
LG pairs.
However, there are three common trends in these simulations. 
First, at most of $1\%$ the pairs are expected to have satellite
distributions with the same asphericity aas the Local Group; second,
between $38\%$ to $59\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $2\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather
typical distribution in the LCDM context, while the MW is atypical in
this respect.  
These results place the LG at the level of a $3\sigma$ outlier
in the LCDM paradigm. 
The approach presented here facilitates the comparison between
asphericity results from different numerical setups and choices to
select halo pairs in simulations, hopefully helping to clarify the
reasons behin the LG atipicallity in LCDM. 
\end{abstract}

@@

Finally, to compute the odds of finding the LG satellite distributions 
in LCDM simulations, we do not make a direct search for the exact
observed values into simulations because, as we show in Section, the
LG is itself rare and we do not have enough simulated systems to
perform such a brute force solution. 
We also avoid this direct comparison for a second reason. 
We do not want to implicitly use either LCDM or observations as what
is considered normal or standard and compute the deviations from that
point. 
Our approach is different. 
We set as a point of reference an spherical satellite
distribution to quantify how both simulations and observations
deviate from the spherical distribution.
Simulations help us to construct an explicit probability distribution
for deviations from sphericity in LCDM.
Then we compute the odds of finding n satellite distributions that
deviates from sphericity as much as the LG does.

@@

Figure \ref{fig:expected_number} summarizes the results from this
experiment. 
The left panel shows the probability density for the number of M31
systems in a parent sampe of $10^4$ pairs, the right panel shows the
results for the MW.
For M31, between $37\%$ to $58\%$ of the pairs have a satellite
distribution as aspherical as M31 in observations; this fraction drops
dramatically for the MW where only $0.03\% to 2\%$ of the satellites
are expected to have extreme aspherical distributions as the MW.

Considering the joint distribution of M31 and MW we find that at most
$1\%$ of the pairs are expected to be similar to the LG.
In a three dimensional gaussian distribution, having a $1\sigma$,
$2\sigma$ and $3\sigma$ interval corresponds to have respectively $19 \%$, $73 \%$ and
$97 \%$ of the total of points in the distribution.
With this result in mind the $LG$ has the same degree of atipicality
as an $3\sigma$ outlier. 
% FUNDAMENTALS OF MULTIVARIATE STATISTICS, for Imaging and optics,
% Bajorski.
%In [150]: stats.chi2.cdf(l**2,3)
%Out[150]: array([ 0.19874804,  0.73853587,  0.97070911])

>>>>

\begin{abstract}
We use numerical simulations to build an explicit probability
distribution for the asphericity in the satellite distribution around
galaxies similar to the Local Group (LG) in the Lambda Cold Dark
Matter (LCDM) paradigm. 
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of resolution, simulated physics and criteria to select
LG pairs.
However, there are three common trends in these simulations. 
First, at most of $2\%$ the pairs are expected to have satellite
distributions with the same asphericity aas the Local Group; second,
between $27\%$ to $56\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $3\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather
typical distribution in the LCDM context, while the MW is atypical in
this respect.  
These results place the LG at the level of a $3\sigma$ outlier
in the LCDM paradigm. 
The approach presented here facilitates the comparison between
asphericity results from different numerical setups and choices to
select halo pairs in simulations, hopefully helping to clarify the
reasons behin the LG atipicallity in LCDM. 
\end{abstract}

@@

Finally, to compute the odds of finding the LG satellite distributions 
in LCDM simulations, we do not make a direct search for the 
observed values into simulations because, as we show in Section, the
LG is itself difficult to fina and we do not have enough simulated systems to
perform such a brute force solution. 
We also avoid this direct comparison because we do not want to
implicitly use either LCDM or observations as what is considered
normal or standard and compute the deviations from that point.

Our approach is different. We set as a point of reference an spherical satellite
distribution to quantify how both simulations and observations
deviate from the spherical distribution, following recent
recommendations presented in \citep{2017AN....338..854P}

Simulations help us to construct an explicit probability distribution
for deviations from sphericity in LCDM.
Then we compute the odds of finding a satellite distribution that
deviates from sphericity as much as the LG does.

@@

Figure \ref{fig:expected_number} summarizes the results from this
experiment. 
The left panel shows the probability density for the number of M31
systems in a parent sampe of $10^4$ pairs, the right panel shows the
results for the MW.
For M31, between $27\%$ to $56\%$ of the pairs have a satellite
distribution as aspherical as M31 in observations; this fraction drops
dramatically for the MW where only $0.02\% to 3\%$ of the satellites
are expected to have extreme aspherical distributions as the MW.

Considering the joint distribution of M31 and MW we find that at most
$2\%$ of the pairs are expected to be similar to the LG.
In a three dimensional gaussian distribution, having a $1\sigma$,
$2\sigma$ and $3\sigma$ interval corresponds to have respectively $19 \%$, $73 \%$ and
$97 \%$ of the total of points in the distribution.
With this result in mind the $LG$ has the same degree of atipicality
as an $3\sigma$ outlier. 
% FUNDAMENTALS OF MULTIVARIATE STATISTICS, for Imaging and optics,
% Bajorski.
%In [150]: stats.chi2.cdf(l**2,3)
%Out[150]: array([ 0.19874804,  0.73853587,  0.97070911])
